Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-20 paper/main.tex
JRM edits to intro

\begin{abstract}
Our moment-by-moment subject experience reflects external sensory information from each moment, along with additional information from our past that we carry with us into that moment.  The blend of memories, knowledge, emotions, goals, and other internal perceptual and mental states that color our subjective experience provides a \textit{context} for interpreting new information and conceptually linking what is happening now with our prior experiences.  Because this contextual information is often person-specific, the subjective experience that each person encodes into their memory is often idiosyncratic, even for shared experiences and sensory perspectives.  We sought to study which aspects of a shared naturalistic experience were preserved or distorted, and how those distortions compared across individuals.  To this end, we developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences.  We model experiences as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration.  We also demonstrate how \textit{memories} may also be modeled as trajectories through the same spaces.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative neuropsychological questions about how we remember naturalistic experiences into quantitive geometric questions about the spatial configurations of trajectory shapes.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectories of participants' recountings of the episode nearly all captured the coarse spatial properties of the original episode's trajectory (i.e., the essential plot points), but participants differed in their memory for fine details.  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains distort and transform our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experience and singularly labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether or not they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of exact recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the \textit{essence} of an experience, and whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) as playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and memories of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or discarded) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: \textbf{1}) the \textit{precision} with which a participant recounts each event, and \textbf{2}) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled'' measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of either the episode or how participants would later recount it.

>>>>

\begin{abstract}
Our ongoing subjective experience reflects external sensory information from each moment, along with additional information from our past that we carry with us into that moment.  The blend of memories, knowledge, emotions, goals, and other internal perceptual and mental states that color our subjective experience provides a \textit{context} for interpreting new information and conceptually linking what is happening now with our prior experiences.  Because this contextual information is often person-specific, the subjective experience that each person encodes into their memory is often idiosyncratic, even for shared experiences and sensory perspectives.  We sought to study which aspects of a shared naturalistic experience were preserved or distorted, and how those distortions compared across individuals.  To this end, we developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences.  We model experiences as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration.  We also demonstrate how \textit{memories} may also be modeled as trajectories through the same spaces.  According to this view, encoding an experience into memory entails geometrically distorting or transforming the original experience's trajectory.  This translates qualitative neuropsychological questions about how we remember naturalistic experiences into quantitive geometric questions about the spatial configurations of trajectory shapes.  We applied our framework to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging.  We found that the trajectories of participants' recountings of the episode nearly all captured the coarse spatial properties of the original episode's trajectory (i.e., the essential plot points), but participants differed in their memory for fine details.  We also identified a network of brain structures that were sensitive to the shape of the episode's trajectory through word embedding space, and an overlapping network that predicted, at the time of encoding, how people would distort (transform) the episode's trajectory when they recounted the episode later.  Our work provides insights into how our brains distort and transform our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experience and singularly labeled as having been recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well-suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of (exact) recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

How might we formally characterize the \textit{essence} of an experience, and whether it has been recovered by the rememberer?  And how might we distinguish an experience's overarching essence from its low-level details?  One approach is to start by considering some fundamental properties of the dynamics of our experiences.  Any given moment of an experience tends to derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details.  The event-level properties reflect how the high-level experience unfolds (i.e., its essence), whereas subtler within-event changes reflect low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) as playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence versus its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and memories of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  The essence of the episode is captured by the sequence of geometric coordinates of its events, and the episode's low-level details are captured by examining its within-event geometric properties.

Comparing the overall shapes of the topic trajectories of the episode and of participants' retellings of the episode reveals which aspects of the episode's essence were preserved (or discarded) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of either the episode or how participants would later recount it.
